sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-11-02 doc/paper-jss/owls.Rmd
Preping JSS

---
title: "Appendix 3: Example of a spatially varying effect of climate on citizen science count data with sdmTMB"
output:
  bookdown::pdf_document2:
    highlight: pygments
    toc: true
    number_sections: true
urlcolor: blue
bibliography: refs.bib
csl: mee.csl
header-includes:
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \linenumbers
---

```{r setup, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.7,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE,
  dev = "png",
  dpi = 150,
  optipng = "-strip all"
)
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
```

# Background

@@

# Preparation of spatial data

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library(ggplot2)
library(patchwork)
library(dplyr)
library(sdmTMB)
library(sf)
```

```{r load-data, cache=FALSE} 
snow <- readRDS(here::here("data/SNOW_data.rds"))
glimpse(snow)
```

@@

```{r mesh, echo=TRUE, out.width="4in"}
mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)

@@

Next, we will plot the owl counts through time to see if there is an obvious overall trend, or relationship with NAO. The dataframe also contains the Southern Oscillation Index (SOI), which is related to El Niño cycles, but we will only consider NAO in this example because any effect of SOI was even weaker.

```{r explore, fig.asp = 0.3, cache=FALSE}
ggplot(snow, aes(year, count, colour = nao)) +
  geom_jitter() +
  scale_colour_viridis_c() +
  theme_light()
```

# Model structure

First, we will try fitting GLMMs (generalized linear mixed effects models) with a few different distributions appropriate for count data. 
<!-- We will use `normalize = TRUE` to speed up optimization.  -->
<!-- Setting `silent = FALSE` would give us optimization details as the model fits but clutter the R Markdown output. -->
Our only fixed effect will be the NAO climate index (`nao`), but because of the wide spatial extent of these surveys, we also expect that the effect of NAO might not be consistent throughout North America. 
Therefore, the effect of NAO will also be allowed to vary spatially using this formula: `spatial_varying = ~ 0 + nao`.
Because there does not appear to be much correlation between years, year will be included as a random factor: `(1 | year_f)`. 
We will also include a fixed spatial random field (`spatial = "on"`) to account for consistent differences in habitat, proximity to the breeding grounds, and effects of flyways and barriers like coasts, and independent and identically distributed spatiotemporal random field (`spatiotemporal = "iid"`) to allow for annual deviations in these spatial effects due to factors like local weather and prey abundance. 
The random intercept for year allows the mean estimate for each of these spatiotemporal fields to vary. 
If years in sequence were strongly correlated with each other, we could use a random walk to link their mean estimates by setting `time_varying = ~ 1`.
It might also be appropriate to include count ID as another random effect `(1 | CBCID)` to account for consistent differences between count circles in habitat and human observers. 
But since its inclusion doesn't change our main result and it slows model fitting, we will rely on the spatial random effects to link observations in the same location for this example.

Since these models will take a while to fit, in practice, we might set `silent = FALSE` to monitor optimization progress. 
However, we will maintain the default `silent = TRUE` to avoid cluttering the document with the output.

```{r rds1, echo=FALSE, cache=FALSE}
rds_saved <- file.exists(here::here("data/snow_w_main_effect_0_150km.rds"))
```

```{r fit-poisson, echo=TRUE, eval=!rds_saved, warning=FALSE}
m0 <- sdmTMB(count ~ 1 + nao + (1 | year_f),
  spatial_varying = ~ 0 + nao,
  time = "year",
  spatial = "on",
  spatiotemporal = "IID",
  family = poisson(link = "log"),
  data = snow, mesh = mesh,
  silent = TRUE
)
```

```{r fit-negb1, echo=TRUE, eval=!rds_saved, warning=FALSE}
m1 <- sdmTMB(count ~ 1 + nao + (1 | year_f),
  spatial_varying = ~ 0 + nao,
  time = "year",
  spatial = "on",
  spatiotemporal = "IID",
  family = nbinom1(link = "log"),
  data = snow, mesh = mesh,
  silent = TRUE
)
```

```{r fit-negb2, echo=TRUE, eval=!rds_saved, warning=FALSE}
m2 <- sdmTMB(count ~ 1 + nao + (1 | year_f),
  spatial_varying = ~ 0 + nao,
  time = "year",
  spatial = "on",
  spatiotemporal = "IID",
  family = nbinom2(link = "log"),
  data = snow, mesh = mesh,
  silent = TRUE
)
```

# Model validation

## Basic inspection

```{r save-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE, eval=TRUE}
if (!rds_saved) {
  saveRDS(m0, here::here("data/snow_w_main_effect_0_150km.rds"))
  saveRDS(m1, here::here("data/snow_w_main_effect_1_150km.rds"))
  saveRDS(m2, here::here("data/snow_w_main_effect_150km.rds"))
}
```

```{r load-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE}
if (rds_saved) {
  m0 <- readRDS(here::here("data/snow_w_main_effect_0_150km.rds"))
  m1 <- readRDS(here::here("data/snow_w_main_effect_1_150km.rds"))
  m2 <- readRDS(here::here("data/snow_w_main_effect_150km.rds"))
}
```

```{r aic, echo=TRUE, cache=FALSE}
AIC(m0, m1, m2)
```

According to AIC, the `nbinom2()` (negative binomial, NB2) distribution has the most parsimonious fit.
We could also check residual distributions at this stage.
We will rerun with `reml = TRUE` to ensure better estimation of the random effects that dominate this model and check that the gradients are all small enough.
If the gradients aren't quite small enough, we can easily run additional optimization steps using the `run_extra_optimization()` function.

```{r sanity, echo=TRUE}
sanity(m2)
```

```{r rds2, echo=FALSE, cache=FALSE}
rds_saved2 <- file.exists(here::here("data/snow_w_main_effect_150km_reml.rds"))
```

```{r fit-negb2-reml, echo=TRUE, eval=!rds_saved2, warning=FALSE, cache=FALSE}
m <- sdmTMB(count ~ 1 + nao + (1 | year_f),
  spatial_varying = ~ 0 + nao,
  time = "year", 
  spatial = "on", 
  spatiotemporal = "iid",
  family = nbinom2(link = "log"),
  reml = TRUE,
  data = snow, mesh = mesh,
  silent = TRUE
)
```

```{r save-reml-model, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE, eval=FALSE}
if (!rds_saved2) {
  saveRDS(m, here::here("data/snow_w_main_effect_150km_reml.rds"))
}
```

@@

```{r check, echo=FALSE, include=FALSE, eval=TRUE, cache=FALSE}
if (rds_saved2) {
  m <- readRDS(here::here("data/snow_w_main_effect_150km_reml.rds"))
  m # printing this first just prevents some excess messages from cluttering the output when printing below
}
```

```{r print-fit, warning=FALSE, echo=TRUE, message=FALSE, cache=FALSE}
print(m)
```

@@

```{r tidy, cache=FALSE}
tidy(m, conf.int = TRUE)
tidy(m, effects = "ran_pars", conf.int = TRUE)
```

@@

```{r predict, message=FALSE}
pred <- predict(m)
set.seed(19208)
pred$resid <- residuals(m)
```

<!-- \clearpage -->

@@

```{r qqnorm, out.width="33%"}
qqnorm(pred$resid)
qqline(pred$resid)
```

More accurate residuals can be generated using tmbstan, but this a slower process.
These are randomized quantile residuals where the fixed effects are fixed at their MLE (maximum likelihoood estimate) values and the random effects are sampled with MCMC via tmbstan/Stan.
To calculate these, we have to use the version of our model fit without REML.
We would ideally sample with more MCMC samples than this and assess convergence by setting `print_stan_model = TRUE`.

```{r stan-resids}
f <- here::here("data/mcmc_residuals.rds") # for caching
set.seed(82919)
if (!file.exists(f)) {
  mcmc_resids <- residuals(m2, type = "mle-mcmc", mcmc_iter = 101, mcmc_warmup = 100)
  saveRDS(mcmc_resids, f)
} else {
  mcmc_resids <- readRDS(f)
}
```

```{r stan-qqnorm, out.width="33%"}
qqnorm(mcmc_resids)
qqline(mcmc_resids)
```

We can also test for zero-inflation using the `simulate.sdmTMB()` function and a couple functions from the DHARMa package.
If our data were zero-inflated or overdispersed, we might have needed to use a delta ("hurdle") model instead (e.g., `delta_truncated_nbinom2()` or possibly even `delta_gamma()` if we were willing to treat the counts as continuous.

```{r zero-test, echo=TRUE, message=FALSE}
s_nb2 <- simulate(m, nsim = 500)
pred_fixed <- m$family$linkinv(predict(m)$est_non_rf)
r_nb2 <- DHARMa::createDHARMa(
  simulatedResponse = s_nb2,
  observedResponse = m$data$count,
  fittedPredictedResponse = pred_fixed
)
DHARMa::testZeroInflation(r_nb2, plot = FALSE)
```

Next, we modify a theme to plot spatiotemporal data and check the distribution of these residuals:

```{r theme, cache=FALSE}
theme_set(theme_bw() + theme(
  plot.title = element_text(size = 12),
  legend.position = "top", legend.justification = c(0, 0), # move legend to top left
  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"),
  strip.background = element_rect(fill = NA, colour = NA), # de-emphasize facet labels
  panel.grid.major = element_line(colour = "grey90"), # for lat/lons
  axis.title = element_blank() # lat/lon is clear from axis text
)) 
```

```{r plot-resids, fig.width=8, dpi=200}
ggplot(pred, aes(X, Y, colour = resid)) +
  geom_point(size = 0.25) +
  facet_wrap(~year) +
  coord_fixed() +
  scale_colour_gradient2(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

# Plotting spatial predictions

To produce pretty maps of these predictions, we back transformed the spatial coordinates into the same units as our original projection and then convert back into an sf object so that the plot axis will be in latitude and longitude.

```{r proj-p}
p <- pred %>% mutate(X = X * 100000, Y = Y * 100000)
p_proj <- p %>% mutate(x = X, y = Y) %>%
  sf::st_as_sf(coords = c("x", "y"), crs = Albers)
```

We retrieved continental coastlines from the package rnaturalearth and outlines of the great lakes from <http://www.naturalearthdata.com/> and transformed them to the correct projection.

@@

We can then compress the annual predictions into mean values for plotting the spatially varying effect of NAO.
Most simply, we can extract the overall coefficients using the `b <- tidy(m)` and add them together with the spatially varying coefficients `zeta_s`.
However, this method does not give us spatially explicit estimates of uncertainty.
To do this, we can get the output from the joint precision matrix simulation draws by using the `predict()` function again, but this time indicating a number of simulation draws (e.g., `sims = 500`) and what component we want estimated (`sims_var = "zeta_s"`). 
We also extract the same number of draws for estimates of the main effect of NAO using the `spread_sims()` function (this time `nsim = 500`). 
The simulation draws for both can then be combined before calculating the median and lower and upper quantiles from these draws (in this case we will use 80%) and adding these values to our prediction data frame.

```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE}
b <- tidy(m, conf.int = TRUE)

nsim <- 500
zeta_s <- predict(m, nsim = nsim, sims_var = "zeta_s")
sims <- spread_sims(m, nsim = nsim)
beta <- sims$nao
combined <- beta + t(zeta_s)

@@

```{r}
nao_effect_lwr95 <- as.numeric(apply(combined, 2, quantile, probs = 0.025))
```

The main effect of NAO indicates that there is a weak average positive effect of annual mean NAO on overall counts `r round(b[b$term == "nao", 2], 2)` (`b[b$term == "nao", 2]`), while for individual count locations the median estimated effects of NAO range from `r round(min(p_mean$nao_sim), 2)` to `r round(max(p_mean$nao_sim),2)` (`range(p_mean$nao_sim)`).
Based on the 80% CI calculated above, the resulting minimum estimated effects of NAO for each count location range from `r round(min(p_mean$nao_lwr), 2)` to `r round(max(p_mean$nao_lwr),2)` (`range(p_mean$nao_lwr)`) and the maximums range from `r round(min(p_mean$nao_upr), 2)` to `r round(max(p_mean$nao_upr),2)` (`range(p_mean$nao_upr)`). 
All these effects are in link space (`link = "log"`) and could be `exp()` to give the expected multiplicative contribution to the predicted owl counts, however the additive nature of the effect in link space is actually somewhat more intuitive to interpret visually, so this is what we will plot on our maps:

```{r plot-map, echo=TRUE, eval=TRUE, out.width="90%", echo=FALSE, cache=FALSE}
ggplot(data = p_proj) +
  geom_point(
    data = p_mean, aes(X, Y, colour = nao_effect, size = mean_est_count), alpha = 0.5
  ) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000), # adjusts space on sides
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(p_mean$nao_lwr), max(p_mean$nao_upr)), 
    guide = guide_colourbar(direction = "horizontal", title.position = "top")
  ) +
  guides(size = "none") +
  labs(colour = "Estimated effect in log space") +
  ggtitle("Map of combined main and spatially varying effects of NAO on Snowy Owl count") +
  theme(legend.position = c(0.1, 0.07), axis.title = element_blank())
```

```{r plot-map-ci, echo=F, eval=TRUE, out.width="90%", cache=FALSE, fig.asp=0.3}
p1 <- ggplot(data = p_proj) +
  geom_point(
    data = p_mean,
    aes(X, Y, colour = nao_lwr, size = mean_est_count), alpha = 0.5
  ) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000), # adjusts space on sides
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(p_mean$nao_lwr), max(p_mean$nao_upr)), 
    guide = guide_colourbar(direction = "horizontal", title.position = "top")
  ) +
  guides(size = "none", colour = "none") +
  # labs(colour = "Lower 90% CI") +
  ggtitle("Lower 80% CI") +
  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())

@@

If we recalculated these at 95% CI the range of minimum estimates would be from `r round(min(nao_effect_lwr95), 2)` to `r round(max(nao_effect_lwr95),2)` meaning that for the most conventional level of statistical confidence, we cannot conclude that the effect of NAO at an annual scale differs from 0. 
However, the southeast to northwest gradient in the intensity illustrated in the maps above is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.
This result suggests that exploring the effects of NAO at finer temporal scales (possibly seasonally) and/or with lagged effects could prove fruitful.

<!-- \clearpage -->

@@

```{r plot-all-effects, echo=TRUE, eval=TRUE, cache=FALSE, dpi=220}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X, Y, colour = exp(est)), size = 0.4, alpha = 0.5) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    trans = "sqrt", # makes variation in low numbers visible
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")
  ) +
  labs(colour = "Predicted counts") +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

<!-- Note that for all of these annually varying predictions, we are showing only the most recent 25 years.  -->

\clearpage
## Other components of predictions

@@

```{r plot-spatial-effects, echo=TRUE, eval=TRUE, out.width="5in", dpi=200}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X, Y, colour = omega_s), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_gradient2(guide = guide_colourbar(
    direction = "horizontal", title.position = "top"
  )) +
  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())
```

\clearpage

@@

Finally, we can look at the spatiotemporal random effects (`epsilon_st`) that represent deviations from the fixed effect predictions and the spatial random effect deviations. 
These represent biotic and abiotic factors causing spatial correlation that are changing through time and are not accounted for in the model. 
For owls, this is most likely to be things like landcover changes, prey abundance cycles, and snow depth.

```{r plot-spatiotemporal-effects, echo=TRUE, eval=TRUE, dpi=210}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = epsilon_st),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_gradient2(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

<!-- Note that for these annually varying predictions, we are showing only the most recent 25 years.  -->

\clearpage

>>>>

```{r setup-owls, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.7,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE,
  dev = "png",
  dpi = 150,
  optipng = "-strip all"
)
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
```

# Example: spatially varying effect on owl counts

## Background

@@

## Preparation of spatial data

```{r packages-owls, message=FALSE, warning=FALSE, cache=FALSE}
library("ggplot2")
library("patchwork")
library("dplyr")
library("sf")
library("sdmTMB")
```

```{r rnaturalearthdata, echo=FALSE}
if (!require("rnaturalearthdata", quietly = TRUE)) {
  stop(
    "Please install 'rnaturalearthdata'.\n",
    "`remotes::install_github('ropensci/rnaturalearthdata')`"
  )
}
```

```{r load-data, cache=FALSE} 
snow <- readRDS(here::here("data/SNOW_data.rds"))
```

Our data frame is called `snow`:

```{r head-snow, echo=TRUE}
head(snow, n = 3)
```

@@

```{r mesh-owls, echo=TRUE, out.width="4in"}
mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)

@@

<!-- Next, we will plot the owl counts through time to see if there is an obvious overall trend, or relationship with NAO. The dataframe also contains the Southern Oscillation Index (SOI), which is related to El Niño cycles, but we will only consider NAO in this example because any effect of SOI was even weaker.  -->

## Model structure

Our only fixed effect will be the NAO climate index (`nao`), but because of the wide spatial extent of these surveys, we also expect that the effect of NAO might not be consistent throughout North America. 
Therefore, the effect of NAO will also be allowed to vary spatially using this formula: `spatial_varying = ~ 0 + nao`.
Because there does not appear to be much correlation between years, year will be included as a random factor: `(1 | year_f)`. 
We will also include a fixed spatial random field (`spatial = "on"`) to account for consistent differences in habitat, proximity to the breeding grounds, and effects of flyways and barriers like coasts, and independent and identically distributed spatiotemporal random field (`spatiotemporal = "iid"`) to allow for annual deviations in these spatial effects due to factors like local weather and prey abundance. 
The random intercept for year allows the mean estimate for each of these spatiotemporal fields to vary. 
If years in sequence were strongly correlated with each other, we could use a random walk to link their mean estimates by setting `time_varying = ~ 1`.
<!-- It might also be appropriate to include count ID as another random effect `(1 | CBCID)` to account for consistent differences between count circles in habitat and human observers.  -->
<!-- But since its inclusion does not change our main result and it slows model fitting, we will rely on the spatial random effects to link observations in the same location for this example. -->
<!-- Since these models will take a while to fit we will set `silent = FALSE` to monitor optimization progress.  -->

We initially fit this model with NB1 and NB2 negative binomial observation error and Poisson error. According to AIC (`AIC()`), the `nbinom2()` (NB2) distribution has the most parsimonious fit. We will therefore proceed with that model.

```{r rds1, echo=FALSE, cache=FALSE}
rds_saved <- file.exists(here::here("data/snow_w_main_effect_0_150km.rds"))
rds_saved2 <- file.exists(here::here("data/snow_w_main_effect_0_150km2.rds"))
```

```{r fit-negb2, echo=TRUE, eval=!rds_saved, warning=FALSE}
fit_owl <- sdmTMB(count ~ 1 + nao + (1 | year_f),
  spatial_varying = ~ nao,
  time = "year",
  data = snow, 
  family = nbinom2(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  mesh = mesh,
  reml = TRUE,
  silent = FALSE
)
```

```{r cache-negb2, echo=FALSE, eval=!rds_saved}
saveRDS(fit_owl, here::here("data/snow_w_main_effect_0_150km.rds"))
```

```{r read-negb2, echo=FALSE, eval=rds_saved}
fit_owl <- readRDS(here::here("data/snow_w_main_effect_0_150km.rds"))
```

## Model validation

We will run the `sanity()` function to apply some basic checks.

```{r sanity, echo=TRUE}
sanity(fit_owl)
```

```
#>   Non-linear minimizer suggests successful convergence
#>   Hessian matrix is positive definite
#>   No extreme or very small eigen values detected
#> x `ln_tau_E` gradient > 0.001
#>   See `?run_extra_optimization()`
#>   Or refit with `control = sdmTMBcontrol(newton_loops = 1)`
#> 
#> x `ln_phi` gradient > 0.001
#>   See `?run_extra_optimization()`
#>   Or refit with `control = sdmTMBcontrol(newton_loops = 1)`
#> 
#>   No fixed-effect standard errors are NA
#>   No fixed-effect standard errors look unreasonably large
#>   No sigma parameters are < 0.01
#>   No sigma parameters are > 100
#>   Range parameter doesn't look unreasonably large
```

We note that some log-likelihood gradients are slightly large for some fixed effects.

```{r}
fit_owl$gradients
```

We can run additional optimization steps using the `run_extra_optimization()` function.

```{r owl-newton, results='hide', eval=!rds_saved2}
fit_owl2 <- run_extra_optimization(fit_owl, nlminb_loops = 0, newton_loops = 1)
```

```{r cache-negb2-2, echo=FALSE, eval=!rds_saved2}
saveRDS(fit_owl, here::here("data/snow_w_main_effect_0_150km2.rds"))
```

```{r read-negb2-2, echo=FALSE, eval=rds_saved2}
fit_owl2 <- readRDS(here::here("data/snow_w_main_effect_0_150km2.rds"))
```

And confirm that the gradients are smaller:

```{r sanity2, echo=TRUE}
sanity(fit_owl2)
```

```
#> Non-linear minimizer suggests successful convergence
#> Hessian matrix is positive definite
#> No extreme or very small eigen values detected
#> No gradients with respect to fixed effects are >= 0.001
#> No fixed-effect standard errors are NA
#> No fixed-effect standard errors look unreasonably large
#> No sigma parameters are < 0.01
#> No sigma parameters are > 100
#> Range parameter doesn't look unreasonably large
```

@@

```{r print-fit-owls, warning=FALSE, echo=TRUE, message=FALSE, cache=FALSE}
print(fit_owl2)
```

@@

```{r tidy-owls, cache=FALSE}
tidy(fit_owl2, conf.int = TRUE)
tidy(fit_owl2, effects = "ran_pars", conf.int = TRUE)
```

@@

```{r predict-owls, message=FALSE}
pred <- predict(fit_owl2)
set.seed(19208)
pred$resid <- residuals(fit_owl2)
```

@@

```{r qqnorm-owls, out.width="60%"}
qqnorm(pred$resid)
qqline(pred$resid)
```

More accurate residuals can be generated using \pkg{tmbstan}, but this a slower process.
These are randomized quantile residuals where the fixed effects are fixed at their MLE (maximum likelihoood estimate) values and the random effects are sampled with MCMC via \pkg{tmbstan}/\pkg{Stan}.
To calculate these, we have to use the version of our model fit without REML.
We would ideally sample with more MCMC samples than this and assess convergence by setting `print_stan_model = TRUE`.
For the purposes of this building this appendix quickly, we will not run this code.

```{r stan-resids-owls, eval=FALSE}
mcmc_resids <- residuals(fit_owl, type = "mle-mcmc", mcmc_iter = 101, mcmc_warmup = 100)
```

We can also test for zero-inflation using the `simulate.sdmTMB()` function and a couple functions from the \pkg{DHARMa} package.
If our data were zero-inflated or overdispersed, we might have needed to use a delta ("hurdle") model instead (e.g., `delta_truncated_nbinom2()` or possibly even `delta_gamma()` if we were willing to treat the counts as continuous.

```{r zero-test-owls, echo=TRUE, message=FALSE}
s_nb2 <- simulate(fit_owl2, nsim = 500)
pred_fixed <- fit_owl2$family$linkinv(pred$est_non_rf)
r_nb2 <- DHARMa::createDHARMa(
  simulatedResponse = s_nb2,
  observedResponse = fit_owl2$data$count,
  fittedPredictedResponse = pred_fixed
)
DHARMa::testZeroInflation(r_nb2, plot = FALSE)
```

We do not find evidence of zero inflation.

```{r theme-owls, cache=FALSE, echo=FALSE}
theme_set(theme_bw() + theme(
  plot.title = element_text(size = 12),
  legend.position = "top", legend.justification = c(0, 0), # move legend to top left
  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"),
  strip.background = element_rect(fill = NA, colour = NA), # de-emphasize facet labels
  panel.grid.major = element_line(colour = "grey90"), # for lat/lons
  axis.title = element_blank() # lat/lon is clear from axis text
)) 
```

## Plotting spatial predictions

To produce maps of these predictions, we back transform the spatial coordinates into the same units as our original projection and then convert back into an \pkg{sf} object so that the plot axes will be in latitude and longitude.

```{r proj-p}
p <- pred %>% mutate(X = X * 100000, Y = Y * 100000)
p_proj <- p %>% mutate(x = X, y = Y) %>%
  sf::st_as_sf(coords = c("x", "y"), crs = Albers)
```
{
Next we retrieve continental coastlines from the package \pkg{rnaturalearth} and outlines of the great lakes from <http://www.naturalearthdata.com/> and transform them to the correct projection.

@@

We can then compress the annual predictions into mean values for plotting the spatially varying effect of NAO.
Most simply, we can extract the overall coefficient using the `b <- tidy(fit_owl2)` and add it to the spatially varying coefficients `zeta_s`.
However, this method does not give us spatially explicit estimates of uncertainty.
To do this, we can get the output from the joint precision matrix simulation draws by using the `predict()` function again, but this time indicating a number of simulation draws (e.g., `nsim = 200`) and what component we want estimated (`sims_var = "zeta_s"`). 
We also extract the same number of draws for estimates of the main effect of NAO using the `spread_sims()` function (this time `nsim = 200`). 
The simulation draws for both can then be combined before calculating the median and lower and upper quantiles from these draws (in this case we will use 80%) and adding these values to our prediction data frame.

```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE}
b <- tidy(fit_owl2, conf.int = TRUE)

nsim <- 200
zeta_s <- predict(fit_owl2, nsim = nsim, sims_var = "zeta_s")
sims <- spread_sims(fit_owl2, nsim = nsim)
beta <- sims$nao
combined <- beta + t(zeta_s)

@@

```{r, echo=FALSE}
nao_effect_lwr95 <- as.numeric(apply(combined, 2, quantile, probs = 0.025))
```

The main effect of NAO indicates that there is a weak average positive effect of annual mean NAO on overall counts `r round(b[b$term == "nao", 2], 2)` (`b[b$term == "nao", 2]`), while for individual count locations the median estimated effects of NAO range from `r round(min(p_mean$nao_sim), 2)` to `r round(max(p_mean$nao_sim),2)` (`range(p_mean$nao_sim)`).
Based on the 80% CI calculated above, the resulting minimum estimated effects of NAO for each count location range from `r round(min(p_mean$nao_lwr), 2)` to `r round(max(p_mean$nao_lwr),2)` (`range(p_mean$nao_lwr)`) and the maximums range from `r round(min(p_mean$nao_upr), 2)` to `r round(max(p_mean$nao_upr),2)` (`range(p_mean$nao_upr)`). 
All these effects are in link space (`link = "log"`) and could be exponentiated to give the expected multiplicative contribution to the predicted owl counts; however, the additive nature of the effect in link space is more intuitive to interpret visually, so this is what we will plot on our maps:

```{r plot-map-owls, echo=FALSE, eval=TRUE, out.width="90%", echo=FALSE}
ggplot(data = p_proj) +
  geom_point(
    data = p_mean, aes(X, Y, colour = nao_effect, size = mean_est_count), alpha = 0.5
  ) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000), # adjusts space on sides
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(p_mean$nao_lwr), max(p_mean$nao_upr)), 
    guide = guide_colourbar(direction = "horizontal", title.position = "top")
  ) +
  guides(size = "none") +
  labs(colour = "Estimated effect in log space") +
  ggtitle("Map of combined main and spatially varying effects of NAO on Snowy Owl count") +
  theme(legend.position = c(0.1, 0.07), axis.title = element_blank())
```

```{r plot-map-ci, echo=FALSE, eval=TRUE, out.width="90%", cache=FALSE, fig.asp=0.3}
p1 <- ggplot(data = p_proj) +
  geom_point(
    data = p_mean,
    aes(X, Y, colour = nao_lwr, size = mean_est_count), alpha = 0.5
  ) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000), # adjusts space on sides
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(p_mean$nao_lwr), max(p_mean$nao_upr)), 
    guide = guide_colourbar(direction = "horizontal", title.position = "top")
  ) +
  guides(size = "none", colour = "none") +
  # labs(colour = "Lower 90% CI") +
  ggtitle("Lower 80% CI") +
  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())

@@

If we recalculated these at 95% confidence intervals levels the range of minimum estimates would be from `r round(min(nao_effect_lwr95), 2)` to `r round(max(nao_effect_lwr95),2)` meaning that for the most conventional level of statistical confidence, we cannot conclude that the effect of NAO at an annual scale differs from 0. 
However, the southeast to northwest gradient in the intensity illustrated in the maps above is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.
This result suggests that exploring the effects of NAO at finer temporal scales (possibly seasonally) and/or with lagged effects could prove fruitful.

@@

```{r plot-all-effects-owls, echo=FALSE, eval=TRUE, cache=FALSE, dpi=220}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X, Y, colour = exp(est)), size = 0.4, alpha = 0.5) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    trans = "sqrt", # makes variation in low numbers visible
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")
  ) +
  labs(colour = "Predicted counts") +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

## Other components of predictions

@@

```{r plot-spatial-effects-owls, echo=FALSE, eval=TRUE, out.width="5in", dpi=200}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X, Y, colour = omega_s), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_gradient2(guide = guide_colourbar(
    direction = "horizontal", title.position = "top"
  )) +
  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())
```

@@

Finally, we can look at the spatiotemporal random effects (`epsilon_st`) that represent deviations from the fixed effect predictions and the spatial random effect deviations. 
These represent biotic and abiotic factors causing spatial correlation that are changing through time and are not accounted for in the model. 
For owls, this is most likely to be factors like landcover changes, prey abundance cycles, and snow depth.

```{r plot-spatiotemporal-effects-owls, echo=FALSE, eval=TRUE, dpi=210}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = epsilon_st),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_gradient2(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```
